---
title: "AspleniaDataExploration"
format: html
editor: visual
---

## Introduction

Here we will explore the asplenia data.

```{r}
#| label: load-packages
library("readxl")
```

## Load data

Using ReadXL to read from Excel spreadsheet.

There are 3095 rows and 823 columns.
That's a lot of data.

```{r}
#| label: load-data
x <- read_xlsx("Khoury_Clean_Asplenia_COVID study_18102023.xlsx")
dim(x)
```


## Look at some simple data visualisation

### Gender

Here we will do a simple analysis of the gender composition of this group.

```{r}
#| label: gender composition

par(mar=c(3,15,2,1))
vals <- table(x$`What is your gender?`)
vals <- vals[order(vals)]
barplot(vals,xlim=c(0,2000),main="gender breakdown",xlab="no. participants",cex.names=0.8,las=1,horiz = TRUE)
text(vals+150,((1:length(vals)-0.4)*1.2),labels = vals)

```

## Asplenia in gender groups

Now we will look at the spleen status in these gender groups.

```{r}
#| label: gender balance by asplenia

par(mar=c(3,15,25,1))

woman <- subset(x,`What is your gender?`=="Woman")
man <- subset(x,`What is your gender?`=="Man")
womanvals <- table(woman$`Have you ever been diagnosed with a spleen that doesn't work OR had it surgically removed?`)
womanvals <- womanvals[order(womanvals)]
names(womanvals) <- paste("(W)",names(womanvals))
manvals <- table(man$`Have you ever been diagnosed with a spleen that doesn't work OR had it surgically removed?`)
manvals <- manvals[order(manvals)]
names(manvals) <- paste("(M)",names(manvals))
catnames <- names(table(x$`Have you ever been diagnosed with a spleen that doesn't work OR had it surgically removed?`))

vals <- cbind("men"=manvals,"women"=womanvals)
par(mar=c(5,5,5,1))
barplot(vals,horiz = TRUE,las=1,main="spleen status by gender",xlab="no. participants",col=c("white","lightgray","darkgray"))
legend("right",legend=catnames,fill = c("white","lightgray","darkgray"),title="spleen status")
text(c(15,70,700,40,160,800),1.2*(c(0.9,1.1,1,1.9,2.1,2)-0.4),labels = vals,col="black",cex=1)

pc <- cbind("man"=manvals/sum(manvals)*100,"woman"=womanvals/sum(womanvals)*100)
barplot(pc,horiz = TRUE,las=1,main="spleen status by gender",xlab="proportion participants (%)",col = c("white","lightgray","darkgray","white","lightgray","darkgray"))
legend("right",legend=catnames,fill = c("white","lightgray","darkgray"),title="spleen status")
pcr <- signif(pc,digits = 2)
text(c(1.5,6,45,2.5,10,45),1.2*(c(0.9,1.1,1,1.9,2.1,2)-0.4),labels = pcr,col="black",cex=1)

#plot(0,type='n',axes=FALSE,ann=FALSE)

legend("right",legend=catnames,fill = c("white","lightgray","darkgray"),title="spleen status")

```

## Session information

```{r}
#| label: session

sessionInfo()

```
